---
title: "Rule-Based Data Binning for Credit Risk Modeling in R"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "Contents"
execute:
  warning: false
  message: false
---

# Introduction to Binning

Binning is a fundamental data preprocessing technique used across various data science, machine learning, and statistical modeling tasks. It involves converting continuous numeric variables into discrete categories or intervals known as "bins." This approach is commonly applied in:

-   **Customer segmentation**

-   **Marketing analytics**

-   **Sensor data analysis**

-   **Medical data grouping**

-   **Financial risk modeling**

## Why Do We Use Binning?

Binning helps address challenges when continuous data is complex, noisy, or difficult to interpret. General benefits include:

-   Converts detailed numeric values into understandable categories—for Example, High, Medium, and Low.

-   Minimizes the influence of extreme values or anomalies.

-   Allows business users and non-technical stakeholders to easily interpret results.

-   It helps break data into meaningful groups for targeted strategies.

-   Missing values can be categorized and analyzed separately.

## Importance of Binning in Risk Analysis

In financial and credit risk analysis, binning is critical because risk scores or predictive model outputs are often continuous but need to be:

-   Explained to non-technical stakeholders.

-   Used for rule-based decision-making (e.g., "Reject if High Risk") and flagging based on Risk analysis.

Grouped to compare default rates or risk levels.

In the **Home Credit Default Risk dataset**, binning external risk scores like EXT_SOURCE_1 helps:

-   Simplify complex numeric data.

-   Segment customers by risk level (Low, Medium, High)

-   Analyze default rates across groups.

-   Handle missing risk scores properly.

# Loading Libraries and Data

## Load Required Libraries {#Load Required Libraries}

```{r}
library(tidyverse)  # for data manipulation
library(readr)      # for reading csv
library(ggplot2)
library(dplyr)
```

## Load Dataset

```{r}
# Replace the file path with your dataset location
data <- read_csv("application_train.csv")

# Preview the data
glimpse(select(data, EXT_SOURCE_1, TARGET))


```

# Binning a Numeric Predictor (EXT_SOURCE_1)

Convert the numeric predictor into categories **Low, Medium, High**, and **Missing** if NA.

```{r}
# Define a binning function
bin_ext_source <- function(x) {
  if (is.na(x)) {
    return("Missing")
  } else if (x <= 0.3) {
    return("Low")
  } else if (x <= 0.7) {
    return("Medium")
  } else {
    return("High")
  }
}

# Apply the function to create a new binned column
data <- data %>%
  mutate(EXT_SOURCE_1_BINNED = sapply(EXT_SOURCE_1, bin_ext_source))

# Check the counts in each bin
table(data$EXT_SOURCE_1_BINNED)

```

-   **High Score Group (28,764 clients)**: These are customers with EXT_SOURCE_1 scores above 0.7. **They are low-risk clients** expected to have a low probability of default.

-   **Low Score Group (27,459 clients)**: Customers with EXT_SOURCE_1 scores less than or equal to 0.3. **These are high-risk clients more likely to default.**

-   **Medium Score Group (77,910 clients)**: Moderate-risk customers with scores between 0.3 and 0.7.

**Missing (173,378 clients)**: A significant number of clients have missing EXT_SOURCE_1 values. This group needs careful handling as missing could:

-   Contain hidden risk

-   Be due to data collection issues

### Importance of this Distribution:

-   The **"Missing" bin dominates** the dataset, raising questions:

    -   Should we treat missing as high-risk or low-risk?

    -   Does missing correlate with default behavior?

-   Proper treatment of this group is crucial in credit scoring models. It can be done using Sensitivity Analysis / Modeling Scenarios:

    **Scenario 1:** Treat Missing as a medium-risk group (keep it).

    **Scenario 2:** Run the model excluding Missing and compare results.

    **Scenario 3:** Impute missing values if external sources are available.

## Visualize Binned Variable

```{r}
library(ggplot2)

ggplot(data, aes(x = EXT_SOURCE_1_BINNED)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Binned EXT_SOURCE_1 Distribution", x = "EXT_SOURCE_1 Bin", y = "Count") +
  theme_minimal()
```

# Default Rate Analysis by Risk Group

## Calculate Default Rates

```{r}
default_rate <- data %>%
  group_by(EXT_SOURCE_1_BINNED) %>%
  summarise(DefaultRate = mean(TARGET, na.rm = TRUE), Count = n())
default_rate
```

-   **High Score Group** (28,764 clients): Low-risk group with the lowest default rate (3%).

-   **Low Score Group** (27,459 clients): High-risk group with the highest default rate (14.5%).

-   **Medium Group** (77,910 clients): Moderate risk with a 6.7% default rate.

-   **Missing Group** (173,378 clients): Default rate of 8.5%, higher than medium and high groups, suggesting some hidden risk.

## Visualize Default Rate

```{r}
ggplot(default_rate, aes(x = EXT_SOURCE_1_BINNED, y = DefaultRate, fill = EXT_SOURCE_1_BINNED)) +
  geom_bar(stat = "identity") +
  labs(title = "Default Rate by EXT_SOURCE_1 Risk Bin", x = "Risk Group", y = "Default Rate") +
  theme_minimal()

```

Key Insight:

-   Clear evidence that lower external scores correlate with higher default risk.

-   Missing group requires special attention due to its size and relatively high default rate.

# Statistical Testing

## Chi-Square Test

```{r}
chisq_test <- chisq.test(table(data$EXT_SOURCE_1_BINNED, data$TARGET))
chisq_test
```

-   The chi-square test is highly significant (p \< 2.2e-16), confirming that default rates differ significantly across the risk bins.

-   Strong statistical evidence that the binning strategy is meaningful in separating risk levels.

# Logistic Regression Using Binned Variable

The logistic regression model is built to:

-   **Quantify the impact** of each risk bin (Low, Medium, High, Missing) on the probability of default.

-   **Validate binning effectiveness** by checking if the groups significantly differ in predicting default.

-   **Support risk-based decision-making** by estimating how much safer or riskier one group is compared to another.

-   **Handle binary outcomes** like default (TARGET=1) or non-default (TARGET=0), which logistic regression is designed for.

## Prepare Data for Modeling

```{r}
data <- data %>% mutate(EXT_SOURCE_1_BINNED = factor(EXT_SOURCE_1_BINNED, levels = c("Low", "Medium", "High", "Missing")))
model_data <- data %>% select(TARGET, EXT_SOURCE_1_BINNED) %>% drop_na()
```

## Build Logistic Regression Model

```{r}
model_binned <- glm(TARGET ~ EXT_SOURCE_1_BINNED, data = model_data, family = "binomial")
summary(model_binned)

```

-   The intercept represents the log-odds of default for the baseline "Low" risk group.

-   **Medium Group**: Shows a significant negative coefficient (-0.86203), meaning less likelihood of default compared to the "Low" risk group.

-   **High Group**: Strong negative coefficient (-1.68106), indicating very low default risk compared to "Low".

-   **Missing Group**: Also less risky than "Low" but riskier than "High" or "Medium".

# Predict Default Probability for a High-Score (Low-Risk of Defaulting) Example

```{r}
new_client <- data.frame(EXT_SOURCE_1_BINNED = factor("High", levels = c("Low", "Medium", "High", "Missing")))
predict(model_binned, newdata = new_client, type = "response")
```

A customer in the "High" score bin has only a 3.05% chance of defaulting, confirming the low-risk nature of this group.

# Key Takeaway

-   Binning makes numeric scores explainable.

-   Low-score groups have significantly higher default risk.

-   Logistic regression confirms that "High" and "Medium" groups are safer.

-   The chi-square test and modeling validate binning as a strong strategy for risk prediction.

-   **Logistic regression quantifies the risk difference** across bins, providing clear, actionable insights.

-   Binned variables improve modeling, decision-making, and communication with non-technical stakeholders.
